---
title: "Bayesian statistics model methodology"
output:
    html_document:
        theme: united
        df_print: kable
date: 'Compiled: `r Sys.Date()`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This vignette highlights the reasoning and methodology used to generate probabilistic qualifying predictions within `laps-of-judgement` via Bayesian hierarchical modelling. 

### Load in the data

Practice session data for the given year and round is collected via `FastF1` using the `python/get_fp_data.py`. You must run this script before attempting to generate qualifying time predictions.

```{r init}
library(brms)
library(cmdstanr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(parallel)
library(tidybayes)
library(patchwork)

# Load the utility functions
source(here::here("R/utils.R"))

# Set the default argument parameters
target_race <- "Spanish Grand Prix"
year <- 2025
max_stint <- 6

# Load in the lap time data
data <- read.csv(glue::glue(here::here("data/processed/all_fp_laps_{year}.csv"))) |>
  filter(RoundName == target_race) |>
  parse_lap_times() |>
  add_elapsed_time()
```

The resulting data frame contains one row per practice lap for the specified round. 

### Filter the practice session data

To improve the accuracy of the qualifying simulation, laps are filtered to retain only those representative of a low-fuel, single-lap 'qualifying' efforts. Laps on used tyres, in-laps, out-laps, and those affected by traffic or yellow flags are removed. A maximum stint length of `max_stint` laps is applied to exclude long run data, and the 107% rule drops any remaining outliers that would distort the prior.

```{r filter}
# Filter for qualifying runs
quali_data <- filter_qualifying_laps(data, max_stint = max_stint)
stopifnot(mean(is.na(data$LapTime_sec)) < 0.1)

# Apply 107% rule
fastest_lap <- min(quali_data$LapTime_sec)
model_data_q <- filter(quali_data, LapTime_sec <= fastest_lap * 1.07)
```

### Bayesian statistics

The initial assumptions (priors) of the Bayesian inference model account for the average lap time: `normal(intercept_prior, 5)` and regularises erratic swings in lap times using the standard deviation and residual error (sigma) of lap times.

Applying a Gaussian likelihood provides a highly robust and computationally efficient approximation for valid push laps, which tend to cluster tightly towards the mean. While the Gaussian model serves as an effective baseline for this statistical inference, future iterations exploring Log-Normal or Ex-Gaussian distributions could theoretically better capture the inherent right-hand skew of the stochastic data.

```{r fit, echo=TRUE, include=TRUE, message=FALSE, warning=FALSE, results='hide'}
# Intercept prior centres the median lap time
intercept_prior <- round(median(model_data_q$LapTime_sec, na.rm = TRUE))

# Fit the Bayesian model
fit_quali <- brm(
  LapTime_sec ~ log(Weekend_Mins_Elapsed + 1) + (1 | Team / Driver),
  data = model_data_q,
  family = gaussian(),
  prior = c(
    prior_string(paste0("normal(", intercept_prior, ", 5)"),
                 class = "Intercept"),
    prior(exponential(1), class = "sd"),
    prior(exponential(1), class = "sigma")
  ),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  cores = detectCores(),
  threads = threading(max(1, floor(detectCores(
  ) / 4))),
  backend = "cmdstanr",
  stan_model_args = list(stanc_options = list("O1"))
)
```

This approach attempts to account for the track evolution over the weekend (`log(Weekend_Mins_Elapsed + 1)`) to pull lap times down as the grip improves. Partial pooling shares information across drivers within their respective constructors: `1 | Team / Driver`, allowing extreme outliers (or reduced running) do not destabilise a driver's predicted qualifying time, instead shrinking it towards the team's mean lap times. The model deploys Markov Chain Monte Carlo (MCMC) algorithms via Stan (`backend = "cmdstanr"`), running 4 parallel chains for 4000 iterations to explore thousands of valid parameter combinations.

The posterior distribution output provides a spectrum of probable lap times for each driver, allowing quantification of the uncertainty and subsequent probabilistic forecast of a 'most likely' lap time per driver.

```{r lap, echo=FALSE, results='hide'}
# Get driver information
driver_teams <- model_data_q |> select(Driver, Team) |> distinct()

# Colour index
team_colours <- read.csv(here::here("data/team_colours.csv"))

# Set new prediction data
new_quali_data <- data.frame(
  Driver = driver_teams$Driver,
  Team = driver_teams$Team,
  Weekend_Mins_Elapsed = max(model_data_q$Weekend_Mins_Elapsed, na.rm = TRUE)
)

# Fit simulated qualifying laps
simulated_quali_laps <- posterior_predict(
  fit_quali,
  newdata = new_quali_data,
  allow_new_levels = TRUE
)

# Build a predicted grid
predicted_grid <- data.frame(
  Driver = new_quali_data$Driver,
  Team = new_quali_data$Team,
  Predicted_Time = apply(simulated_quali_laps, 2, quantile, probs = 0.05)
) |>
  arrange(Predicted_Time) |>
  mutate(Predicted_Grid_Position = row_number())

# Apply confidence checks
low_confidence <- as.character(
  model_data_q |>
    group_by(Driver) |>
    filter(all(confidence == "*")) |>
    distinct(Driver) |>
    pull(Driver)
)
predicted_grid$Confidence <- ""
predicted_grid <- predicted_grid |>
  mutate(Confidence = ifelse(Driver %in% low_confidence, "*", Confidence))

# Reformat the labels
predicted_grid <- predicted_grid |>
  mutate(
    Gap = Predicted_Time - min(Predicted_Time),
    Gap_label = ifelse(
      Predicted_Grid_Position == 1,
      sprintf("%d:%06.3f",
              floor(min(Predicted_Time) / 60), min(Predicted_Time) %% 60),
      sprintf("+%.3f", Gap)
    )
  ) |>
  left_join(team_colours |> select(Team, Colour), by = "Team")
```

```{r plots, echo=FALSE, fig.align='center', fig.dpi=300, fig.height=10, results='asis', out.width='100%'}
# View posterior distributions
posterior_forecasts <- new_quali_data %>%
  add_predicted_draws(fit_quali, ndraws = 1000)

posterior_forecasts <- posterior_forecasts |>
  left_join(team_colours |> select(Team, Colour), by = "Team")

# Posterior forecast prediction distribution plot
p1 <- ggplot(posterior_forecasts, aes(x = .prediction, y = reorder(Driver, -.prediction))) +
  stat_halfeye(aes(fill = Colour),
    .width = c(0.66, 0.95),
    point_interval = median_qi
  ) +
  scale_fill_identity() +
  scale_colour_identity() +
  labs(
    title = paste(year, target_race),
    subtitle = "Proabalistic Forecast Distribution",
    x = "Predicted Lap Time (seconds)"
  ) +
  ylab("Driver") +
  theme_minimal()

# Standard worfklow plot
p2 <- ggplot(predicted_grid,
       aes(x = Gap, y = reorder(Driver, -Predicted_Grid_Position))) +
  geom_col(aes(fill = Colour), width = 0.65, colour = NA) +
  geom_text(
    aes(
      label = Gap_label,
      x = ifelse(Gap > 0.35, Gap - 0.01, Gap + 0.01),
      hjust = ifelse(Gap > 0.35, 1, 0),
      colour = ifelse(Gap > 0.35, "white", "black")
    ),
    family = "mono",
    size = 3.1,
    fontface = "bold"
  ) +
  geom_text(
    aes(
      label = Confidence,
      x = ifelse(Gap > 0.35, Gap + 0.05, Gap + 0.15),
      hjust = ifelse(Gap > 0.35, 1, 0),
      colour = "black"
    ),
    family = "mono",
    size = 3.1,
    fontface = "bold"
  ) +
  scale_fill_identity() +
  scale_colour_identity() +
  labs(
    title = paste(year, target_race),
    subtitle = "Predicted Qualifying Gaps",
    caption = if (any(predicted_grid$Confidence == "*"))
      "* low confidence prediction"
    else
      NULL
  ) +
  ylab("Driver") +
  theme_minimal()

p1 | p2
```